### 2017-05-09 Paper_v1.tex
Apr9

%\usepackage{algorithm2e}
%\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}
\hyphenation{dec-ades} %Latex does not know how to hyphenate this by default, so we add it to the list

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described all of the training data, especially the central carbon metabolism.
The model predicted CAT production with a carbon yield equal to 16\%, and an energy efficiency equal to 9\%, of that of a physiologically realistic case calculated using sequence-specific flux balance analysis.
This suggests that CAT production could be further optimized.
The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

In this study, we developed an ensemble of kinetic cell-free protein synthesis (CFPS) models using dynamic metabolite measurements in an \textit{E. coli} cell free extract.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
Characteristic values for model parameters and initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
The ensemble of parameter sets described the training data with a median cost that was greater than two orders of magnitude smaller than random sets constructed using the literature parameter constraints. We then used the ensemble of kinetic models to analyze the CFPS reaction.
First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network and glucose consumption.
CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis. The CAT yield estimated from the kinetic model was equal to 16\% of the theoretical yield when physiologically realistic constraints were used.
Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

% In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
% The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
% This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
% The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
% Using this approach, we have built an ensemble of models that predict an experimental dataset for CAT production in a cell-free \textit{E. coli} system.
% We calculated the CAT yield for the ensemble at 45\% of the theoretical maximum according to constraint-based modeling, and determined through sensitivity analysis that CAT production was most sensitive to the parameters associated with CAT synthesis and the synthesis of GTP, GMP, and amino acids.
% While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
% In the second case (ii), there was a high flux through fumarate dehydrogenase from aspartic acid uptake, whereas in cases i and iii, acetate and lactate accumulation occurred.
% However, oxidative phosphorylation is not operating at its highest efficiency, since there is a mixture of acetate and lactate accumulation.
% In addition, a handful of alanine synthesis pairwise sensitivities were also large.

@@

To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:CAT_balances}).
The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
This shows that there is much room for improvement of the efficiency of CFPS.
Gene knockouts in the electron transport chain further reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on CAT production and carbon yield.
A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
reducing the CAT carbon yield to 6.9\%.
Knocking out \textit{app} increased CAT yield to 8.1\%, but this increase was not statistically significantly different from that of the control.
Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not completely eliminate the production of CAT.
This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
%The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.

Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:oxKO_ssFBA_yields}).
Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
We chose this time window (during the first phase of CAT production) because it was approximately linear both in glucose consumption and in the accumulation of most organic acids.
As the ssFBA calculation had no adjustable parameters, bounds on transcription and translation rates and biochemical fluxes were either estimated from data or from mechanistic models parameterized from literature.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
by sampling plausible ranges for these parameters.
The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:oxKO_ssFBA_yields}C).
Thus, the molecular description of transcription and translation were consistent with experimental measurements.
Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints;
(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E. coli} extract preparation.
(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E. coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}D).
%The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
%For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
%Lastly, when metabolite constraints were applied with experimental measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
The physiological theoretical maximum CAT carbon yield (case i) was 49.3\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
For case ii, the optimal CAT carbon yield was 48.9\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
Lastly, when metabolite constraints based on experimental measurements were applied (case iii), the predicted carbon yield was 6.4\% $\pm$ 2.9\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
Unsurprisingly, this range of carbon yield encompasses both the best-fit set (7.9\%) and the experimental dataset (8.2\%).
For cases i and ii the energy efficiencies were 72.1\% $\pm$ 9.5\% and 71.2\% $\pm$ 9.6\%, respectively, while for case iii it was only 5.1\% $\pm$ 2.4\%.
This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
This range of energy efficiency encompasses both the best-fit set and the experimental dataset (7\%).
However, the model and the experimental dataset fall short of the optimum (cases i and ii).
This suggests there is much room for improvement, primarily by reducing byproduct formation.
%This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
%Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.

To investigate the differences in carbon yield and energy efficiency, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases (Fig.~\ref{fig:NetworkFlux}).
%All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
In cases i and ii, glycolysis was used for energy generation, and most of the carbon flux accumulated as acetate.
The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
Case iii heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued primarily through the Entner–Doudoroff pathway toward pyruvate.
Case iii also predicted the accumulation of pyruvate, lacatate, acetate, and carbon dioxide, which contributed to the much lower carbon yield and energy efficiency.
In all cases, the energy source was primarily oxidative phosphorylation converting cofactors generated throughout the network into ATP.
%Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow.
Taken together, this suggested CAT production could be increased by reducing the accumulation of acetate, lactate, and carbon dioxide.

@@

In this study we present an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
CFPS is seen to be biphasic relying on glucose during the first hour and pyruvate and lactate afterward.
Allosteric control was essential to the maintenance of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
Sensitivity analysis showed that certain substrates and energy species are instrumental to CAT production and overall metabolism.
The system heavily relied on oxidative phosphorylation for the system's energetic needs as well as for CAT synthesis.
A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state, showing its crucial role in CFPS.
In comparing flux distributions between low and high yield cases, carbon flux could be potentially diverted toward CAT by reducing acetate overflow and minimizing flux through the Entner-Doudoroff pathway.
Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.

We present an ensemble of models that quantitatively describes the system behavior of cell-free metabolism and production of CAT.
Experimental observations of the metabolites and cometabolites validate the structure of the model and the estimation of kinetic parameters.
This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
Thus, supplementation with additional GTP may improve the efficiency of CAT production.
A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
Nevertheless, there are further reactions with considerable impact on model performance.
In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
%Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
%The next most sensitive reaction is the conversion of NADPH to NADH, which helps to supply NADH to oxidative phosphorylation reactions.
%Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.

%The ensemble of models captured the dynamic distribution of metabolic flux.
%The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
%while the constraint-based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
% We present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
% While the early models of Shuler and coworkers achieved large-scale, dynamic descriptions of single cells \citep{1984_domach_shuler_BiotechBioeng_01}, and the stoichiometric models associated with the FBA approach are computationally efficient amd widespread \citep{2009_feist_palsson_NatRevMicrobio}, none have yet been able to construct genome-scale kinetic models of cell-free \textit{E. coli} metabolism.
% This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
%However, the low sensitivities to glucose and oxygen only apply in the specific case of the parameter sets that were studied.
%When other species' initial conditions and associated parameters are different, glucose and oxygen could become more important to CAT production.

A logical next step for this work would be sequence-specific dynamic modeling, as the kinetic modeling approach in this study used a single reaction to approximate CAT synthesis.
Including specific transcription and translation steps for CAT would allow more accurate modeling of the complexity and the resource cost of protein synthesis.
Sensitivity analysis could then be performed on these new parameters to determine the robustness of CAT synthesis to the processes of transcription and translation.
%Another area for future work is to more thoroughly sample parameter space.
%Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
%In a different region of parameter space, the system may behave differently but still fit the experimental data.
%This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
%Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
In addition, further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
%how important they are to protein synthesis, and under which conditions.
Finally, the approach should be extended to other protein products.
CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
An important extension of this study would be to apply its insights to other protein applications, where possible.

@@

\section*{Author's contributions}
J.V directed the modeling study.
K.C and J.S conducted the cell free protein synthesis experiments.
J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
J.V and M.V performed the sequence-specific flux balance analysis calculations.
The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.

@@

% Figures and captions go here ...
% \begin{figure}[ht]
% \centering
% \includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
% \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
% \label{fig:Carbon}
% \end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue or gray shaded region) over the ensemble of 100 sets.}
\label{fig:Carbon}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
\caption{Energy species and energy totals by base in the presence of allosteric control. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
\label{fig:Energy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Amino.pdf}
\caption{Amino acids in the presence of allosteric control. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
\label{fig:Amino}
\end{figure}

>>>>

%\usepackage{algorithm2e}
%\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}
\hyphenation{dec-ades} %Latex does not know how to hyphenate this by default, so we add it to the list
\def\wl{\par \vspace{\baselineskip}}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described all of the training data, especially the central carbon metabolism.
XXX Talk about carbon and energy balances XXX
This suggests that CAT production could be further optimized.
The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

In this study, we developed an ensemble of kinetic cell-free protein synthesis (CFPS) models using dynamic metabolite measurements in an \textit{E. coli} cell free extract.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
Characteristic values for model parameters and initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
The ensemble of parameter sets described the training data with a median cost that was greater than two orders of magnitude smaller than random sets constructed using the literature parameter constraints. We then used the ensemble of kinetic models to analyze the CFPS reaction.
First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network and glucose consumption.
CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:CAT_balances}).
The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
This shows that there is much room for improvement of the efficiency of CFPS.
A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.

@@

In this study we present an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
CFPS is seen to be biphasic relying on glucose during the first hour and pyruvate and lactate afterward.
Allosteric control was essential to the maintenance of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
Sensitivity analysis showed that certain substrates and energy species are instrumental to CAT production and overall metabolism.
The system heavily relied on oxidative phosphorylation for the system's energetic needs as well as for CAT synthesis.
A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state, showing its crucial role in CFPS.
Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.

We present an ensemble of models that quantitatively describes the system behavior of cell-free metabolism and production of CAT.
Experimental observations of the metabolites and cometabolites validate the structure of the model and the estimation of kinetic parameters.
This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
Thus, supplementation with additional GTP may improve the efficiency of CAT production.
A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
Nevertheless, there are further reactions with considerable impact on model performance.
In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.

%Another area for future work is to more thoroughly sample parameter space.
%Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
%In a different region of parameter space, the system may behave differently but still fit the experimental data.
%This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
%Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
In addition, further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
Finally, the approach should be extended to other protein products.
CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
An important extension of this study would be to apply its insights to other protein applications, where possible.

@@

\section*{Author's contributions}
J.V directed the modeling study.
K.C and J.S conducted the cell free protein synthesis experiments.
J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.

@@

% Figures and captions go here ...

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Fig_1_Carbon.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue or gray shaded region) over the ensemble of 100 sets.}
\label{fig:Carbon}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Fig_2_Energy.pdf}
\caption{Energy species and energy totals by base in the presence of allosteric control. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
\label{fig:Energy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Fig_3_Amino.pdf}
\caption{Amino acids in the presence of allosteric control. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
\label{fig:Amino}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Fig_4_EnsembleVsRandom.pdf}
\caption{Log of cost function across 37 datasets for data-trained ensemble (blue) and randomly generated ensemble (red, gray background). Median (bars), interquartile range (boxes), range excluding outliers (dashed lines), and outliers (circles) for each dataset. Median across all datasets (large bar overlaid).}
\label{fig:BoxPlot}
\end{figure}
